---
title: "Project1: Wind power energy"
output: 
  html_document:
  toc: true
  toc_float: true
  # toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(magrittr)

```


# Data
The original data is organized in six variables - the first three of them describing the date, the fourth describing the average 
daily wind power production, the fifth and sixth describing the wind speed in m/s and wind direction in radians respectively. 

Before working with this dataset the three variables concerning the date were merged into one variable containing all three. 
Firstly, a graphical presentation of data were created and can be seen as (...) below. It is clearly seen tat none of the variables of the data 
is normally distributed. The *wind speed* looks rather rightly skewed, and the *wind direction* looks left skrewed or could possibly be a 
bimodal distribution. 


*** present some summary statistics.

Before we move on to the modelling of the data, data transformations will be performed. 

(Before you start the model-ling it will make some of the questions simpler if
you normalize (to numbers strictly between 0 and 1) the power production.
Be careful to do this in a meaningful way (hint: relate to installed capacity)
and describe precisely how it is done.)


## Load
```{r}
wind_data <- read.table("../Data/tuno.txt", header = T, sep = " ")
summary(wind_data)
```

## Plot

```{r}
# ggpairs(~pow.obs + ws30 + wd30, wind_data)
pairplot  <- ggpairs(wind_data, columns = c("pow.obs", "ws30", "wd30"))
wind_data$r.day
pairplot
ggsave("Plots/pairplot_wind.png", pairplot)
```

qq-plots herunder

```{r}
qqnorm(wind_data[pow.obs], ylim, main = "Normal Q-Q Plot",
       xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
       plot.it = TRUE, datax = FALSE, …)
qqline(wind_data[1])

plot(wind_data$r.day)
wind_data$r.day %>% as.factor() %>%  density() %>% plot()
```



# Data transformations

## Box-Cox
```{r}
bc.trans <- function(lambda,y){
    y.l <- (y^lambda-1)/lambda
    if(lambda==0){y.l <- log(y)}
    return(y.l)
}
```


## Log-transformation
```{r}
#Log
```



# Models


## Normal distribution
```{r}

```

## Multimodal distribution
```{r}

```


## Von Mises distribution
```{r}

```


## Gamma distribution 
```{r}

```







