---
title: "Project1: Wind power energy part 3"
output: 
  html_document:
  toc: true
  toc_float: true
  # toc_collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Packages

```{r include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(magrittr)
library(MASS)
```


## Load
```{r}
wind_data <- read.delim("../Data/tuno.txt", header = T, sep = " ")
summary(wind_data)
wind_data$pow.obs.norm <- (wind_data$pow.obs)/5000
wind_data$ws30.norm <- (wind_data$ws30)/max(wind_data$ws30)

wind_data

#normalized_pow.obs <- scale(wind_data$pow.obs)
#colMeans(normalized_pow.obs)
#apply(normalized_pow.obs, 2, sd)
#var(wind_data$pow.obs)
#var(wind_data$ws30)
#var(wind_data$wd30)

```






```{r}
data_transform2 <- function(y, lambda){
  2 * log(y^lambda/((1-y)^(1-lambda)))
}
data_transform1 <- function(y, lambda){
  1/lambda*log((y^lambda)/(1-y^lambda))
}
wind_data$pow.lambda1 <- data_transform1(wind_data$pow.obs.norm, 0.2)


model = lm(pow.lambda1 ~ ws30 + I(ws30^2), data = wind_data)
summary(model)
ws = seq(0,25, 1e-3)
df = data.frame(ws30 = ws, ws30_2 = ws^2)
plot(wind_data$ws30, wind_data$pow.lambda1)

lines(ws, predict(model, newdata = df, interval = "confidence")[,1], lty = 2)

```

# Fit parameters

```{r}
res = model$residuals
var(res)
res_matrix <- function(res){
  n = length(res) - 1
  M = matrix(nrow = n, ncol = 2)
  
  for (i in 1:n){
    M[i, 1] = res[i]
    M[i, 2] = res[i+1]
  }
  return(M)
}

nll_multivariate<- function(par, data){
  sigma = par[1]
  par[2] -> p
  Sigma = matrix(data = c(sigma, p*sigma, p*sigma, sigma), nrow = 2, ncol = 2, byrow = T)
  n = length(res_matrix(res))/2
  sum = 0
  
  for (i in 1:n){
      x <- data[i,]
      sum = sum - 1/2*(log(det(Sigma)) + t(x) %*% solve(Sigma) %*% x + 2*log(2*pi))
  
  }
  return(-sum)
}

opt <- nlminb(start = c(1,.5), nll_multivariate, lower = c(1e-16,1e-16), upper = c(200,0.999), data = res_matrix(res))


opt
```

### 
```{r}
hess <- numDeriv::hessian(nll_multivariate, opt$par, data = res_matrix(res))
CI_sigma <- opt$par[1]+c(-1,1)*sqrt(solve(hess[1,1]))*qnorm(0.975)
cat("The wald confidence interval for sigma^2 is ", CI_sigma)
CI_rho = opt$par[2]+c(-1,1)*sqrt(solve(hess[2,2]))*qnorm(0.975)
cat("The wald confidence interval for rho is ", CI_rho)
```


# Compare fisher information

Algebraic fsiher information from a bivariate normal
$$
I\left(\sigma^{2}, \rho\right)=\left(\begin{array}{cc}\frac{n}{\sigma^{4}} & -\frac{n \rho}{\sigma^{2}\left(1-\rho^{2}\right)} \\ -\frac{n \rho}{\sigma^{2}\left(1-\rho^{2}\right)} & \frac{n\left(1+\rho^{2}\right)}{\left(1-\rho^{2}\right)^{2}}\end{array}\right)
$$


```{r}
cat("Numerical matrix values: ", hess)
sigma_MLE <- opt$par[1]
rho_MLE <- opt$par[2]
n <- length(res)

I1 <- n/sigma_MLE^2
I2 <- -n*rho_MLE/(sigma_MLE*(1-rho_MLE^2))
I4 <- n*(1+rho_MLE^2)/(1-rho_MLE^2)^2

cat("\nAlgebraic matrix values: ", matrix(c(I1, I2, I2, I4), nrow = 2, byrow = T))
```

## CHANGE VALUES!!!
$$
\begin{align}
I_{num}\left(\hat{\sigma^2}, \hat{\rho}\right) &= \begin{bmatrix} 374.9 &&  -111.9 \\ -111.9 && 384.0 \end{bmatrix} \\
I_{alg}\left(\hat{\sigma^2}, \hat{\rho}\right) &= \begin{bmatrix} 376.2 && -112.3 \\ -112.3 && 385.3 \end{bmatrix}
\end{align}
$$
Wow, numerical methods results in nearly the same as the algebraic.
# Contour Plots
```{r}
sigma = seq(10,13, by = 0.1)
p = seq(0.15,0.45, by = 0.01)

data = res_matrix(res)
print(length(sigma))
Z = matrix(data = NA, nrow = length(sigma), ncol = length(p))
for (i in 1:length(sigma)){
  print(i)
  for (j in 1:length(p)){
    
    Z[i,j] = nll_multivariate(c(sigma[i], p[j]), data)

  }
  
}
A = -Z -(max(-Z))
confint = c(.99, .975, .95, .9, .8, .5)
C = exp(-1/2*qchisq(confint, df = 1))
contour(x = sigma, y = p, exp(A), level=C, xlab = expression(sigma^2), ylab = expression(rho), main = "Countour Plot of confidence regions", labels = confint)
```






# P values for rho. Wald and Likelihood test
## Likelihood ratio test
```{r}
nll_multivariate2<- function(sigma, p, data){
  Sigma = matrix(data = c(sigma, p*sigma, p*sigma, sigma), nrow = 2, ncol = 2, byrow = T)
  n = length(res_matrix(res))/2
  sum = 0
  
  for (i in 1:n){
      x <- data[i,]
      sum = sum - 1/2*(log(det(Sigma)) + t(x) %*% solve(Sigma) %*% x + 2*log(2*pi))
  
  }
  return(-sum)
}



Profile_rho <- function(rho){
  sigma = nlminb(start = c(1), nll_multivariate2, lower = c(1e-16), upper = c(20), p = rho, data = res_matrix(res))$par
  
  nll_multivariate2(sigma, rho, data = res_matrix(res))


}


opt_rho = nlminb(start = c(0.5), Profile_rho, lower = c(1e-16), upper = c(0.999))

objective = -opt_rho$objective
h0 = -Profile_rho(0)

Q = -2*(h0-objective)
p = pchisq(Q, df = 1, lower.tail = F)
cat("P-value for likelihood ratio test:", p, "\n")
se = sqrt(solve(hess[2,2]))

z = (objective-h0)/se
p1 = 1-pnorm(z)

cat("P-values for Wald test:", p1)


```

## Profile likelihood and quadratic aprroximation
```{r}
rho <- seq(0,0.5, by = 0.01)
L = sapply(rho,Profile_rho)
plot(rho, -L-max(-L), 'l', ylab = "Normalised Log Likelihood", xlab = expression(rho))
C = -1/2*qchisq(1-0.05, df = 1)
abline(h = C)
curvature = 1/solve(hess)[2,2]
lines(rho,-1/2*curvature * (rho - opt_rho$par)^2, 'l', lty = 2)
```




### Z transform stuff profile likelihood ... idunno ... 
```{r}


Z_transform <- function(p){
  1/2 * log((1+p)/(1-p))
}


curvature = solve(hess)[2,2]
SE = sqrt(curvature)
SE_Z <- SE * numDeriv::grad(Z_transform, opt_rho$par)
curvature_Z = 1/(SE_Z^2)
plot(Z_transform(rho), -L-max(-L), 'l')
# lines(rho, -L-max(-L), lty = 2)
lines(Z_transform(rho), -1/2*curvature_Z * (Z_transform(rho) - Z_transform(opt_rho$par))^2, 'l', lty = 2)
print("Vaaary niccce")

```
```{r}
nll_multivariate_transform<- function(par_T, data){
  sigma = exp(par_T[1])
  p = (exp(2*par_T[2])-1)/(exp(2*par_T[2])+1)
  Sigma = matrix(data = c(sigma, p*sigma, p*sigma, sigma), nrow = 2, ncol = 2, byrow = T)
  n = length(res_matrix(res))/2
  sum = 0
  
  for (i in 1:n){
      x <- data[i,]
      sum = sum - 1/2*(log(det(Sigma)) + t(x) %*% solve(Sigma) %*% x + 2*log(2*pi))
  
  }
  return(-sum)
}



opt_trans = nlminb(start = c(0.5, 0.5), nll_multivariate_transform, data = res_matrix(res))
c(log(opt$par[1]), Z_transform(opt$par[2]))

hess_Z <- numDeriv::hessian(nll_multivariate_transform, opt_trans$par, data = res_matrix(res))

1/solve(hess_Z)[2,2]

sigma_MLE <- log(opt$par[1])
rho_MLE <- Z_transform(opt$par[2])
n <- length(res)

I1 <- n/(sigma_MLE^2)
I2 <- -n*rho_MLE/(sigma_MLE*(1-rho_MLE^2))
I4 <- n*(1+rho_MLE^2)/(1-rho_MLE^2)^2

cat("\nAlgebraic matrix values: ", matrix(c(I1, I2, I2, I4), nrow = 2, byrow = T))

H = matrix(c(I1, I2, I2, I4), nrow = 2, byrow = T)


```



```{r}
df2 = cbind(wind_data %>% pull(ws30),(wind_data %>% pull(ws30))^2)
ARreg = arima(wind_data %>% pull(pow.lambda1), order = c(1,0,0), xreg = df2 )
summary(model)
ARreg

```



