---
title: 'Project3: Financial data part 3'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r Packages, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(magrittr)
```


# Data

## Load
```{r}
finance_data <- read.csv("../Data/finance_data.csv", header = T, sep = ";") %>% 
  # tidyr::separate("time", c("year", "month", "day"), sep = "-") %>% 
  dplyr::mutate(time = lubridate::ymd(time)) 
  
summary(finance_data)

#Looking at the data
hist(finance_data$SLV, breaks = 50)
plot(finance_data$time, finance_data$SLV, type="l")
acf(finance_data$SLV) #Alt undtagen lag nr. 16 er inden for confint. 

```

## Mixture models 

### a) Fit normal mixture models with 2 and 3 components

Here we are using the code form lecture 10, where Jan is fitting a poission mixture model with 2 components. We are modifying his code a bit due to the poission having one parameter and the normal dist having two, the mean and the standard deviation.

```{r}
## Normal mixture: transform
## Natural to working parameters
norm_mix_pn2pw <- function(m, mu, sigma, delta) {
  if(sum(delta) >= 1) {
    print("sum(delta) should be < 1")
    return()
  }
  t_sigma <- log(sigma)
  t_delta <- log(delta/(1 - sum(delta))) #t_ angiver working parameters
  return(list(mu = mu, t_sigma = t_sigma, t_delta = t_delta))
}

## Working to natural parameters
norm_mix_pw2pn <- function(m, mu, t_sigma, t_delta){
  if(m == 1){
    return(exp(t_sigma))
  }
  sigma <- exp(t_sigma)
  delta <- exp(t_delta)/(1 + sum(exp(t_delta)))
  delta <- c(1 - sum(delta),delta)
  return(list(mu = mu, sigma = sigma, delta = delta))
}

## Negative log likelihood
nll <- function(theta, m=2, x=finance_data$SLV){
  if(m == 1) {
    return(-sum(pnorm(x, theta[1], exp(theta[2]), log=TRUE))) 
  }
  mu <- theta[1:m]
  t_sigma <- theta[(m+1):(2*m)]
  t_delta <- theta[(2*m+1):(3*m-1)]
  n_pars <- norm_mix_pw2pn(m, mu, t_sigma, t_delta)
  n <- length(x)
  nll <- 0
  for(i in 1:n) {
    nll <- nll - log(sum(n_pars$delta * dnorm(x[i], mu, n_pars$sigma)))
  }
  return(nll)
}


```


### 2 components

```{r}
## Estimation with 2 distributions
m <- 2; 

## Initial values
mu <- mean(finance_data$SLV)*c(1/2,3/2)
sigma <-sd(finance_data$SLV)*c(1/2,3/2)
delta <- c(1/2)

## Working parameters
wpars2 <- norm_mix_pn2pw(m, mu, sigma, delta)
theta2 <- c(wpars2$mu, wpars2$t_sigma, wpars2$t_delta)

## MLE
opt2 <- nlminb(theta2, nll, m = m, x = finance_data$SLV)

## Natural parameters
npars2 <- norm_mix_pw2pn(m, opt2$par[1:m], opt2$par[(m+1):(2*m)], opt2$par[(2*m+1):(3*m-1)])

npars2 
```

```{r}
mix.dist <- function(x ,npars){
  sum(npars$delta * dnorm(x, mean = npars$mu, sd = npars$sigma))
}

## Plot
par(mfrow=c(1,1))
hist(finance_data$SLV, prob=TRUE, nclass=60)
lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), sapply(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), mix.dist, npars=npars2), col=2)

lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars2$delta[1]*dnorm(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars2$mu[1], npars2$sigma[1]), col=4)

lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars2$delta[2]*dnorm(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars2$mu[2], npars2$sigma[2]), col=4)
legend("topleft",  c("2 components",'Components seperately'), col=c("red",'blue'), lty=1, cex=0.7)




```



### 3 Components

```{r}
## Estimation with 3 normal distributions
m <- 3;

## Initial values 
mu <- mean(finance_data$SLV)*c(1/2,1,3/2)
sigma <- sd(finance_data$SLV)*c(1/2,1,3/2);
delta <- c(1/3,1/3)

## Working parameters
wpars <- norm_mix_pn2pw(m, mu, sigma, delta)
theta <- c(wpars$mu, wpars$t_sigma, wpars$t_delta)

## MLE
opt3 <-nlminb(theta, nll, m = m, x = finance_data$SLV)

## Natural parameters
npars3 <- norm_mix_pw2pn(m, opt3$par[1:m], opt3$par[(m+1):(2*m)], opt3$par[(2*m+1):(3*m-1)])

```


```{r}
mix.dist <- function(x ,npars){
  sum(npars$delta * dnorm(x, mean = npars$mu, sd = npars$sigma))
}

## Plot
par(mfrow=c(1,1))
hist(finance_data$SLV, prob=TRUE, nclass=60)
lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), sapply(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), mix.dist, npars=npars3), col=2)

lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars3$delta[1]*dnorm(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars3$mu[1], npars3$sigma[1]), col=4)

lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars3$delta[2]*dnorm(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars3$mu[2], npars3$sigma[2]), col=4)

lines(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars3$delta[3]*dnorm(seq(min(finance_data$SLV), max(finance_data$SLV), 0.001), npars3$mu[3], npars3$sigma[3]), col=4)
legend("topleft",  c("3 components",'Components seperately'), col=c("red",'blue'), lty=1, cex=0.7)


```



## Check of the models

Now we will take a look at the modelsby their AIC and the deviance. 

```{r}
## AIC
AIC <- 2*c(opt2$objective, opt3$objective) + 2*c(length(opt2$par), length(opt3$par))
AIC #-1489.644 -1484.256

## Deviance 
1-pchisq(-2*(opt3$objective-opt2$objective),df=length(opt3$par)-length(opt2$par)) 
#0.8937244
```


We see, that the model with two parameters are slightly better. When we compaire to the best model found in assignment 1, the best is the t-distribution with an AIC = -1490, hence this is the overall best model.





### b) Report confidence interval for the parameters in the best mixture model. 

```{r}
library(numDeriv)
H <- hessian(nll, opt2$par)
se<- sqrt(diag(solve(H)))
(CI1 <- opt2$par[1]+c(-1,1)*se[1]*qnorm(0.975))
(CI2<- opt2$par[2]+c(-1,1)*se[1]*qnorm(0.975))
#[1] -0.0007570256  0.0086514630
#[1] -0.02985318 -0.02044469
```
The mean value is centered around zero. An since, sigma is small, little variance in the data is expected. 



## c) Make a profile likelihood plot of one of the variance parameters in the two component model.  

```{r}

#fra Tobias 

## Profile likelihood for sigma 1 given working parameters
lp.sigma1 <- function(sigma1, m, x, pars0){
  ## Fun for inner optim
  fun.tmp <- function(theta, sigma1, m, x){
    pars <- c(theta[1:m], log(sigma1), theta[-(1:m)])
    nll(pars, m, x)
  }
  nlminb(pars0, fun.tmp, sigma1 = sigma1, m = m, x = x)$objective    
}

## Estimation with 2 distributions
m <- 2; 

## Initial values
mu <- mean(finance_data$SLV)*c(1/2,3/2)
sigma <-sd(finance_data$SLV)*c(1/2,3/2)
delta <- c(0.1)

## Working parameters
wpars <- norm_mix_pn2pw(m, mu, sigma, delta)
theta0 <- c(wpars$mu, wpars$t_sigma, wpars$t_delta)
theta <- c(theta0[1:m],theta0[(m+2):(3*m-1)])

sigma1 <- seq(0.01, 0.1, length=100)

## profile likeihood
pnll <- sapply(sigma1, lp.sigma1, m = m, x = finance_data$SLV, pars0 = theta)

## Plot the profile likelihood
plot(sigma1,exp(-(pnll-min(pnll))),type="l", ylim=c(0,1))
lines(range(sigma1),
      c(1,1)*exp(-qchisq(0.95,df=1)/2),col=2,lty=2,lwd=2)
rug(npars2$sigma,col=2,lwd=2)






```



## d) Reparametricering 
```{r}

## Repara Negative log likelihood

## Profile likelihood for sigma 1 given working parameters
re.lp.sigma1 <- function(sigma1, m, x, pars0){
  ## Fun for inner optim
  fun.tmp <- function(theta, sigma1, m, x){
    pars <- c(theta[1:m], log(sigma1), theta[-(1:m)])
    nll(pars, m, x)
  }
  nlminb(pars0, fun.tmp, sigma1 = sigma1, m = m, x = x)$objective    
}

## Estimation with 2 distributions
m <- 2; 

## Initial values
mu <- mean(finance_data$SLV)*c(1/2,3/2)
sigma <-sd(finance_data$SLV)*c(1/2,3/2)
delta <- c(0.1)

## Working parameters
wpars <- norm_mix_pn2pw(m, mu, sigma, delta)
theta0 <- c(wpars$mu, wpars$t_sigma, wpars$t_delta)
theta <- c(theta0[1:m],theta0[(m+2):(3*m-1)])

sigma1 <- seq(0.01, 0.1, length=100)

## profile likeihood
pnll <- sapply(sigma1, lp.sigma1, m = m, x = finance_data$SLV, pars0 = theta)

## Plot the profile likelihood
plot(sigma1,exp(-(pnll-min(pnll))),type="l", ylim=c(0,1))
lines(range(sigma1),
      c(1,1)*exp(-qchisq(0.95,df=1)/2),col=2,lty=2,lwd=2)
rug(npars2$sigma,col=2,lwd=2)








## Reprametrizing mix-model 

## Natural to working parameters
re.norm.mix.pn2pw <- function(m, mu, sigma, delta) {
  if(sum(delta) >= 1) {
    print("sum(delta) should be < 1")
    return()
  }
  t.sigma <- log(c(sigma[1],sigma[-1]-sigma[-m])) #sigma 2 er nu blevet til differensen mellem sigma1 og sigma2
  t.delta <- log(delta/(1 - sum(delta)))
  return(list(mu = mu, t.sigma = t.sigma, t.delta = t.delta))
}

## Working to natural parameters
re.norm.mix.pw2pn <- function(m, mu, t.sigma, t.delta){
  if(m == 1){
    return(exp(t.sigma))
  }
  sigma <- cumsum(exp(t.sigma)) #lægger sigma1 tilbage til sigma2, så det ikke længere er en differens
  delta <- exp(t.delta)/(1 + sum(exp(t.delta)))
  delta <- c(1 - sum(delta),delta)
  return(list(mu = mu, sigma = sigma, delta = delta))
}

## Negative log likelihood
re.nll <- function(theta, m, x){
  if(m == 1) {
    return(-sum(pnorm(x, theta[1], exp(theta[2]), log=TRUE))) 
  }
  mu <- theta[1:m]
  t.sigma <- theta[(m+1):(2*m)]
  t.delta <- theta[(2*m+1):(3*m-1)]
  n.pars <- re.norm.mix.pw2pn(m, mu, t.sigma, t.delta)
  n <- length(x)
  nll <- 0
  for(i in 1:n) {
    nll <- nll - log(sum(n.pars$delta * dnorm(x[i], mu, n.pars$sigma)))
  }
  return(nll)
}

#########################################################################################################################

## Profile likelihood for sigma 1 given working parameters
re.lp.sigma1 <- function(sigma1, m, x, pars0){
  ## Fun for inner optim
  fun.tmp <- function(theta, sigma1, m, x){
    pars <- c(theta[1:m], log(sigma1), theta[-(1:m)])
    re.nll(pars, m, x)
  }
  nlminb(pars0, fun.tmp, sigma1 = sigma1, m = m, x = x)$objective    
}

m <- 2; 

## Initial values
mu <- mean(finance_data$SLV)*c(1/2,3/2)
sigma <-sd(finance_data$SLV)*c(1/2,3/2)
delta <- c(0.1)

## Working parameters
wpars <- re.norm.mix.pn2pw(m, mu, sigma, delta)
theta0 <- c(wpars$mu, wpars$t.sigma, wpars$t.delta)
theta <- c(theta0[1:m],theta0[(m+2):(3*m-1)])

sigma1 <- seq(0.01, 0.05, length=100)

## profile likeihood
pnll <- sapply(sigma1, re.lp.sigma1, m = m, x = finance_data$SLV, pars0 = theta)

## Plot the profile likelihood
plot(sigma1,exp(-(pnll-min(pnll))),type="l", ylim=c(0,1))
lines(range(sigma1),
      c(1,1)*exp(-qchisq(0.95,df=1)/2),col=2,lty=2,lwd=2)
rug(npars2$sigma[1],col=2,lwd=2)






```

   

## Markov Chain
```{r}
y <- finance_data$SLV

```

```{r}
source("A1.R")
y <- finance_data$SLV

## 2 - state 
## Initial values
m <- 2
lambda0 <- quantile(y,c(0.25,0.75))
gamma0 <- matrix(0.05,ncol=m,nrow=m)
diag(gamma0) <- 1-(m-1)*gamma0[1,1]


## optimize
fit2 <- pois.HMM.mle.nlminb(y,m,lambda0,gamma0)
fit2
# virker ikke sådan rigtigt. der er en masse non-integer probs...


## 3 - state  Denne er ikke god...
## Initial values
m <- 3
lambda0 <- quantile(y,c(0.25,1,0.75))
gamma0 <- matrix(0.05,ncol=m,nrow=m)
diag(gamma0) <- 1-(m-1)*gamma0[1,1]


## optimize
fit3 <- pois.HMM.mllk(y,m,lambda0,gamma0)
fit3

```

```{r}
## working parameters
parvect  <- pois.HMM.pn2pw(m,fit3$lambda,fit3$gamma)
## Optimize (hessian = TRUE return hessian)
mod <- nlm(pois.HMM.mllk,parvect,x=y,m=m,
            hessian=TRUE)  
mod

## Organize the result
parvect <- mod$estimate
names(parvect) <- c("lambda1","lambda2","lambda3","tau21",
                    "tau31","tau12","tau32","tau13","tau23")

se <- sqrt(diag(solve(mod$hessian)))

## Working pars + standard error
round(cbind(parvect,se),digits=2) ## note se of tau31
fit3$gamma
```






